A demonstration of the \LaTeXe\ class file for \itshape{SAGE Publications}

### 2015-12-04 body/efort.tex
robust equilibrium

\section{From a guide trajectory to a discrete sequence of contact configurations (Stage 2)}
\label{sec:efort}
As an input of this stage, we consider a truly feasible guide trajectory $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the root of the robot $R$. We now consider the second problem of computing a trajectory $\mathbf{q}^{\overline{0}}(t)$ for the limbs of the robot. Since we assume true feasibility, we know that such a trajectory exists. Contrary to previous works \cite{DBLP:conf/iser/EscandeKMG08,Bouyarmane2009}, the goal here is not to find any such trajectory but rather to select one with particular properties. Specifically, we show here how to build a contact sequence with a small number of contact variations and  good-efficiency and naturalness of the postures.

More precisely, any mapping $\pi$ introduced in \eqref{eq:pi} can be used to expand $\mathbf{q}^0(t)$ into a whole-body trajectory. We propose here a particular construction of $\pi$ leading to interesting contact sequences.

%We compute the image of a mapping $f$ along $\mathbf{q}^0(t)$ into a discrete sequence of contacts $ \in C_{contact}$. We first describe the discretization of $\mathbf{q}^0(t)$ and the algorithm used to create or break contacts along it. Then we describe how contacts are explicitely generated along the discretized trajectory.

\subsection{Extension of the guide trajectory} 
The guide trajectory $\mathbf{q}^0(t)$ is first discretized into a sequence of $j$ key placements:  
\begin{equation*}
	\mathbf{Q}^0 = [\mathbf{q}^0_{0}; \mathbf{q}^0_{i}; ..., \mathbf{q}^0_{j-1}]
\end{equation*} 
where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ respectively correspond to the start and goal configurations.  To ensure continuity in the contact transition phases, we rewrite $\pi$ under the following recursive form for any $0<i<j$:
\begin{equation*}
    \pi\colon\left\{
    \begin{aligned}		
        \mathbf{Q}^0 \in C_{reach} & \longrightarrow C_{contact} \\
        %\mathbf{q}^{0}_0 &  \longrightarrow  \mathbf{q}_{start} \\
        \mathbf{q}^{0}_i &  \longrightarrow  \mathbf{q}^{0}_i \oplus g(\mathbf{q}_{i - 1},\mathbf{q}^{0}_i) 
    \end{aligned}
    \right.
\end{equation*} 
We initialize the recurrence with $\pi(\mathbf{q}^0_{0}) = \mathbf{q}_0$ the initial configuration of the robot.
The function $g$ is defined independently by $g^k$ for each limb $R^k$. In defining $g^k$, two aspects must be considered. Is the limb $R^k$ in contact? And which criteria is it optimizing?

\medskip%\subsection{Maintaining and breaking contact in the sequence}
\noindent\textbf{Maintaining a contact:} If possible, a limb in contact at time $i-1$ remains in contact at $i$. The contact is broken if an inverse kinematics solver fails to find a collision free limb configuration which satisfies joint limits.
The solver is a direct implementation of the inverse kinematics solver with task priorities framework proposed in \cite{Baerlocher2004}, which
handles explicitely joint limits, and allows to integrate collision avoidance as a secondary criterion.

If the solver fails, the contact is broken and a collision free configuration is assigned to the limb.
%~ The configuration is selected to be the closest to the previous configuration (by minimizing $\| \mathbf{p}^k(\mathbf{q_i}) - \mathbf{p}^k(\mathbf{q_{i-1}}) \|$).

Once a first candidate configuration is taken for all limbs, the quasi-static balance is tested by whether the weight wrench is in the gravito-inertial cone (i.e. there exists valid contact forces that compensate for the weight of the robot), using the geometric approach described in \cite{qiu:dhm:2011}. If the balance is not obtained, new contacts are randomly generated using the following procedure.

%\begin{itemize} 
%\item If the robot is not statically balanced and $\mathbf{q}_{i-1}^k$ is not in contact, $g$ computes a configuration $\mathbf{q}_{i}^k$ so that a contact is taken with the environment (Fig.~\ref{fig:generate_contact}).
%\item Else, $g$ computes a configuration $\mathbf{q}_{i}^k$ minimizing $\| \mathbf{p}^k(\mathbf{q_i}) - \mathbf{p}^k(\mathbf{q_{i-1}}) \|$, subject to $R^k(\mathbf{q_i}) \cap O = \emptyset$ and $\mathbf{q_i}^k$ respects the  joint limits of $R^k$.
%
%If $\mathbf{q}_{i - 1}^k$ is in contact, this results in trying to maintain the contact for $\mathbf{q}_{i}^k$ (Figure~\ref{fig:break_contact}--1 and 2). The minimum is different than zero if $W^k(\mathbf{q}_i^0) \cap  \mathbf{p}^k(\mathbf{q_{i-1}}) = \emptyset$ (Figure~\ref{fig:break_contact}--3), or collision can not be  prevented (Figure~\ref{fig:break_contact}--4).
%\end{itemize}

\begin{figure}[t]
\centering
  \begin{overpic}[width=0.6\linewidth]{figures/break_contact}
		\put (3,15) {1)} 
		\put (27,15) {2)} 
		\put (50,15) {3)} 
		\put (75,15) {4)} 
		%~ \put (68,58) {3.a)} 
		%~ \put (5,27) {3.b)} 
		%~ \put (37,27) {4.a)} 
		%~ \put (68,27) {4.b)} 
	\end{overpic}
\caption{Contacts are maintained unless their position is too far, or the environment prevents it.}
		   \label{fig:break_contact}
\end{figure}

\begin{figure}[t]
\centering
  \begin{overpic}[width=0.6\linewidth]{figures/generate_contact}
		%~ \put (5,58) {1)} 
		%~ \put (37,58) {2)} 
		%~ \put (68,58) {3.a)} 
		%~ \put (5,27) {3.b)} 
		%~ \put (37,27) {4.a)} 
		%~ \put (68,27) {4.b)} 
	\end{overpic}
\caption{Contacts are generated when the configuration is not balanced.}
		   \label{fig:generate_contact}
\end{figure}

%\subsection{Generation of a contact configuration}  
\medskip\noindent\textbf{Creating a contact:} We consider a configuration where some limbs are in contact, some are free and quasi-static balance is not enforced. To enforce balance, we proceed in the following 
manner:
we randomly select a contact free limb; if there is no contact free limb, we select
the limb that made contact first.
Using the contact generator introduced in~\cite{Tonneau2014}, we project the configuration of this limb into a contact that enhances balance, if it exists (Figure~\ref{fig:generate_contact}); If balance is not achievable and a contact is possible, it is generated anyway;
If balance is not achieved, the next limb is selected and projected into a contact configuration, and so on.
This approach can lead to the repositioning of existing contacts, in which case intermediate states are inserted to reposition the contacts.
This current implementation does not guarantee that the planner will succeed in generating a balanced configuration, because
true feasibility is not fully guaranteed.
However in practice the planner is successful in the large majority of cases, as discussed in section~\ref{sec:stat1}.
%~ A new contact configuration is obtained by projecting a random configuration to the nearest contacts. The sampling is biased by selecting configurations so that the limbs previously in contact are already close to contact and at least one additional limb is close to the environment. The random configuration is then projected onto the contact by iteratively minimizing the distance between the end effectors and the environment, under the constraint of respecting the joint limits and avoiding the collision.

\subsection{Contribution to the global movement: the EFORT criteria}
\noindent\textbf{EFORT criterion:} If only relying on the random sampling to select new contacts, the planner produces inefficient postures. The resulting contact sequence is then poorly efficient and unnatural. Moreover, the limbs are not well configured and are not able to efficiently follow the general movement: contacts break frequently.

When creating additional contacts, we therefore propose to select particular configurations that allow to exert a force compatible with the direction of motion. This task efficiency is measured  with the Extended FORce Transmission ratio (EFORT)~\cite{Tonneau2014}.
%
The measure of EFORT is given by
\begin{equation}
\alpha_{EFORT}(\mathbf{q}^k, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)\mathbf{m}]^{-\frac{1}{2}} ( \nu_0 \mathbf{n}^T \mathbf{m})
\end{equation}
where $\mathbf{J}$ is the Jacobian matrix of the limb $R^k$ in configuration $\mathbf{q}^k$; $\nu_0$ is the friction coefficient of the contact surface; $\mathbf{n}$ is the normal of the contact surface; and $\mathbf{m}$ is the direction opposite to the motion,
given by the 3D vector connecting $\mathbf{q}_{i}^0$ and $\mathbf{q}_{i+1}^0$.
%
The first part of the equation measures the ratio between the joint  torques and the resulting force applied along $\mathbf{m}$. The second part quantifies the odds of slipping while applying a force along $\mathbf{m}$.

\noindent\textbf{Optimization at creation:} In practice, a database of configurations is stored for each limb, which can be considered as manipulator arms. The database is implemented  as an octree data structure, indexed by the end-effector positions of the configurations (and additionally storing $\mathbf{J}$). 
%Fig.~\ref{fig:contact_gen}-2 presents a few configurations for the right arm of the robot.  
Upon request, the octree returns a set of configurations close to contact (Fig.~\ref{fig:contact_gen}-3). These candidates are sorted based on their task efficiency, given by $\alpha_{EFORT}$. The first candidate in this list satisfying the balance criterion and is collision free is selected and projected on the contact surface using our inverse kinematics solver.

%Therefore to generate a contact, we want to find a configuration which is collision free, allows to maintain balance, and maximizes EFORT. We choose this configuration among a set of possible candidates, rapidly computed thanks to a contact generator, also proposed in~\cite{Tonneau2014}. We recall the overall idea here. We consider each limb of the robot as a manipulator arm attached to the root. We store a database of configurations for each manipulator. The database is implemented  as an octree data structure, indexed by the end-effector positions of the configurations. The jacobian matrix $\mathbf{J}$ of each configuration is also stored. Figure~\ref{fig:contact_gen} -- 2 presents a few configurations for the right arm of the robot.

%To summarize: given a feasible trajectory for the root of the robot $R$, we are able to generate a discrete sequence of contact configurations  describing a feasible motion for $R$, between a start an goal configurations. The contact configurations are chosen to be statically balanced and task efficient, which means that they can generate a force  contributing to the motion of the root along the trajectory.

>>>>

\subsection{Manipulability-based heuristics for locally optimal contact selection}
The other kind of heuristics we use are based on the analysis of the motion capabilities
provided by a candidate configuration regarding the task being achieved.
These heuristics allow to locally optimize a contact location regarding a particular task.
For instance, once can be interested in configurations that allow to efficienctly exert a force compatible with the global direction of motion,
a high velocity in a given direction, or to stay away from singular configurations.
In these section, three such heuristics are derived in this section from the work on manipulability by \cite{Yoshikawa1984}, that we present first.

\subsubsection{The force and velocity ellipsoids.}
For a limb configuration $\mathbf{q}^k$, the jacobian matrix
 $\mathbf{J}(\mathbf{q}^k)$   defines the relation:

\begin{equation} \label{jac}
\mathbf{\dot{p}^k} = \mathbf{J}(\mathbf{q}^k) \dot{\mathbf{q}^k}
\end{equation}

For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}(\mathbf{q}^k)$ as $\mathbf{J}$.

As a linear approximation of the forward kinematics function $f$ , $\mathbf{J}$ describes how small
variations from the configuration $\mathbf{q}$ affect the position vector $\mathbf{p}$.
It can be seen as the transformation mapping a velocity in the configuration space into a velocity in
the euclidian space.

Now we consider the unit ball in the configuration space $C$ defined by the set of joint velocities
for which the norm is inferior or equal to 1:
\begin{equation} \label{ball}
||\dot{\mathbf{q}}||^2 \leq 1
\end{equation}

From Equation~\ref{jac} we can obtain the following equality:
\begin{equation} \label{jaceq}
\mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} = \dot{\mathbf{q}}^T \dot{\mathbf{q}}
\end{equation}

We can use Equation~\ref{jaceq} to map the ball into an ellipsoid in the euclidian space $\mathbb{R}^m$:

\begin{equation} \label{ellipsoid}
\mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} \leq 1
\end{equation}
This ellipsoid is called the manipulability ellipsoid, or velocity ellipsoid, introduced by \cite{Yoshikawa1984}. It describes the set of velocities that can
be reached under the constraints of Equation~\ref{ball} for the current configuration.
The longer an axis of the ellipsoid is, the faster the configuration can move along the direction of the axis.
Figure~\ref{sec:efort_ellipsoid} - left shows the velocity ellipsoid for different configurations of a manipulator with two degrees of freedom.

\begin{figure}[htb]
  \centering
	\begin{overpic}[width=1\linewidth]{figures/EFORT/ellipsoid}
		\put (8.5,1.2) {\small{Velocity ellipsoid}}
		\put (62.5,1.2) {\small{Force ellipsoid} \tiny{(scale 0.5)}}
	\end{overpic}
  % \includegraphics[width=1\linewidth]{ressources/screens/sampling/canap/bd2_cmyk}
  \caption{Examples of velocity and force ellipsoids for a manipulator composed of 2 dofs and 2 segments.
Only the horizontal and vertical speeds are shown (not the rotation speeds), since it would require being able to draw in four dimensions.}
		   \label{sec:efort_ellipsoid}
\end{figure}

Similarly to the velocity ellipsoid, \citeauthor{Yoshikawa1984} also defines the force ellipsoid.
Considering: a force vector $\mathbf{f}$ expressed in the task space $\mathbb{R}^m$;
the equivalent joint torque vector $\mathbf{\tau}$;
we can define the mechanical work in both spaces:
\begin{equation*} \label{power}
\dot{\mathbf{q}}^T \mathbf{\tau} = \dot{\mathbf{p}}^T \mathbf{f}
\end{equation*}

Deriving from Equation~\ref{jac}, we define that the set of achievable force in $\mathbb{R}^m$  subject to the constraint:
\begin{equation*} \label{ballforce}
||\mathbf{\tau}||^2 \leq 1
\end{equation*}
is the force ellipsoid (Figure~\ref{sec:efort_ellipsoid} - right) defined by:
\begin{equation} \label{ellipsoidforce}
\mathbf{f}^T (\mathbf{J}\mathbf{J}^T) \mathbf{f} \leq 1
\end{equation}

\subsubsection{Manipulability based heuristics.}
From these definitions, we can derive three useful heuristics, that all account from the environment and the task being performed.
The first one, EFORT, is introduced by \cite{Tonneau2014}; the other two are new minor contributions, derived from these previous works.

With EFORT, we define the efficiency of a configuration as the ability a limb has to exert a force in a given direction.
We thus consider the force ellipsoid as a basis for our heuristic.
In a given direction $\mathbf{m}$, the length of the ellispoid is given by the force transmission ratio \citep{1087795}:
\begin{equation*}
f_\mathsf{T}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(JJ^{T})\mathbf{m}]^{-\frac{1}{2}}
\end{equation*}

In our problem, to compare candidate configurations, we include the quality of the contact surface, and choose $\mathbf{m}$ as the direction
opposite to the local motion (thus given by the difference between two consecutive root positions):

\begin{equation}
h_{EFORT}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)\mathbf{m}]^{-\frac{1}{2}} ( \mu_0 \mathbf{n}^T \mathbf{m})
\end{equation}
where $\mu_0$ and $\mathbf{n}$ are respectively the friction coefficient and the normal vector of the contact surface.

If the ability to apply important velocities at the effector is considered, we define a new heuristic $h_{vel}$ with a similar reasonning on the velocity ellispoid:
\begin{equation}
h_{vel}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{m}]^{-\frac{1}{2}} ( \mu_0 \mathbf{n}^T \mathbf{m})
\end{equation}

$h_{EFORT}$ and $h_{vel}$ will favor contacts that allow important efforts or fast modifications in the velocity.
$EFORT$ in particular is useful for tasks such as standing up, pushing / pulling.
In other less demanding cases, manipulability can also be considered regarding its original purpose, that means to avoid singularties.
To do so, we can consider the manipulability measure $w$, also given by \citeauthor{Yoshikawa1984}:

\begin{equation*} \label{ellipsoid}
w(\mathbf{q}) = \sqrt{\mathbf{J}\mathbf{J}^T}
\end{equation*}
$w$ measures the ``distance" between a given configuration and a singular configuration. When $w$ is equal to 0, the configuration is in a singular state;
the greater $w$ is, the further away the configuration is from singularity.
We then derive:
\begin{equation}
h_{w}(\mathbf{q}, \mathbf{m}) = \sqrt{\mathbf{J}\mathbf{J}^T} ( \mu_0 \mathbf{n}^T \mathbf{m})
\end{equation}

$h_{EFORT}$, $h_{vel}$ and $h_{w}$ define three kinematic heuristics, fast to compute (indeed, thanks to our sampling based approach, the jacobian an inverse products
can be precomputed offline), that allow the planner to select the more relevant
candidates regarding the local motion task along the guide path.
This result in faster computation times, and a reduced number of contact switches overall.

### 2015-12-04 body/equilibrium.tex
robust equilibrium

\subsection{A heuristic for robust quasi static balance}
\noindent\textbf{EFORT criterion:} If only relying on the random sampling to select new contacts, the planner produces inefficient postures. The resulting contact sequence is then poorly efficient and unnatural. Moreover, the limbs are not well configured and are not able to efficiently follow the general movement: contacts break frequently.

When creating additional contacts, we therefore propose to select particular configurations that allow to exert a force compatible with the direction of motion. This task efficiency is measured  with the Extended FORce Transmission ratio (EFORT)~\cite{Tonneau2014}.
%
The measure of EFORT is given by
\begin{equation}
\alpha_{EFORT}(\mathbf{q}^k, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)\mathbf{m}]^{-\frac{1}{2}} ( \nu_0 \mathbf{n}^T \mathbf{m})
\end{equation}
where $\mathbf{J}$ is the Jacobian matrix of the limb $R^k$ in configuration $\mathbf{q}^k$; $\nu_0$ is the friction coefficient of the contact surface; $\mathbf{n}$ is the normal of the contact surface; and $\mathbf{m}$ is the direction opposite to the motion,
given by the 3D vector connecting $\mathbf{q}_{i}^0$ and $\mathbf{q}_{i+1}^0$.
%
The first part of the equation measures the ratio between the joint  torques and the resulting force applied along $\mathbf{m}$. The second part quantifies the odds of slipping while applying a force along $\mathbf{m}$.

\noindent\textbf{Optimization at creation:} In practice, a database of configurations is stored for each limb, which can be considered as manipulator arms. The database is implemented  as an octree data structure, indexed by the end-effector positions of the configurations (and additionally storing $\mathbf{J}$). 
%Fig.~\ref{fig:contact_gen}-2 presents a few configurations for the right arm of the robot.  
Upon request, the octree returns a set of configurations close to contact (Fig.~\ref{fig:contact_gen}-3). These candidates are sorted based on their task efficiency, given by $\alpha_{EFORT}$. The first candidate in this list satisfying the balance criterion and is collision free is selected and projected on the contact surface using our inverse kinematics solver.

%Therefore to generate a contact, we want to find a configuration which is collision free, allows to maintain balance, and maximizes EFORT. We choose this configuration among a set of possible candidates, rapidly computed thanks to a contact generator, also proposed in~\cite{Tonneau2014}. We recall the overall idea here. We consider each limb of the robot as a manipulator arm attached to the root. We store a database of configurations for each manipulator. The database is implemented  as an octree data structure, indexed by the end-effector positions of the configurations. The jacobian matrix $\mathbf{J}$ of each configuration is also stored. Figure~\ref{fig:contact_gen} -- 2 presents a few configurations for the right arm of the robot.

%To summarize: given a feasible trajectory for the root of the robot $R$, we are able to generate a discrete sequence of contact configurations  describing a feasible motion for $R$, between a start an goal configurations. The contact configurations are chosen to be statically balanced and task efficient, which means that they can generate a force  contributing to the motion of the root along the trajectory.

>>>>

\subsection{A heuristic for robust quasi static balance}
The planner is designed so that any contact configuration it outputs is quasi statically balanced.
Balance is of course critical in multipedal locomotion. For this reason we are interested in a robust 
criterion, that ensures that the robot remains balanced in a real world application, regardless of the errors
that can result from peception and control uncertainties.

We first give a classic linear program (LP) that asserts that a contact configuration is statically balanced.
From this formulation we derive a new LP that returns a quantification of the robustness of the configuration regarding balance.
From this value we can in turn either define a heuristic to choose the more robust candidate, or set a treshold on the required balance robustness
for any configuration. While the LP we propose is new, it is based on an analysis of the problem we propose in \cite{Prete2016}, where the interested reader
can find more details.

\subsubsection{Conditions for static balance.}
We first define the variables of the problem, for a robot $R$ and $e$ contact points, expressed in world coordinates:
\begin{itemize}
\item $\mathbf{c} \in \mathbb{R}^3$ is the robot center of mass (COM);
\item $\mathbf{L}  \in \mathbb{R}^3 $ is the angular momentum at the COM;
\item $\mathbf{m} \in \mathbb{R}$ is the robot mass;
\item $\mathbf{g} = [0,0,-9.81]^T$ is the gravity acceleration;
\item $\mu$ is the friction coefficient;
\item for the i-th contact $1 \leq i \leq e$:
	\begin{itemize}
	\item $\mathbf{p_i}$ is the contact position;
	\item $\mathbf{f_i}$ is the force applied at $\mathbf{p_i}$;
	\item $\mathbf{n}_i,\mathbf{t}_{i1},\mathbf{t}_{i2}$ form a local cartesian coordinate system centered at $\mathbf{p_i}$. $\mathbf{n}_i$ is aligned
	with the contact surface normal, and the $\mathbf{t}_i$s are tangent vectors.
	\end{itemize}
\end{itemize}

According to Coulomb's law, the robot is statically balanced if all contact forces lie in the friction cone defined by the surface.
As classically done, we linearize the friction cone in a conservative fashion with a pyramid, described by four generating rays. We choose for instance:
\begin{equation*}
\mathbf{V}_{i} = \mat{\mathbf{n}_{i} + \mu \mathbf{t}_{i1} & \mathbf{n}_{i} -\mu \mathbf{t}_{i1} & \mathbf{n}_{i} + \mu \mathbf{t}_{i2} & \mathbf{n}_{i} - \mu \mathbf{t}_{i2}}^T
\end{equation*}

Any force belonging to the linearized cone
can thus be express as a linear combination of its four generating rays.
Therefore a constraint on the $\mathbf{f}_i$ is given by the relation:

\begin{equation*}
\forall i  \qquad  \exists \bm{\beta}_i \in \mathbb{R}^{+4} , \mathbf{f}_{i} = \mathbf{V}_{i} \bm{\beta}_i \qquad 
\end{equation*}

we can stack the constraints to obtain:
\begin{equation}\label{eq:gen}
\exists \bm{\beta} \in \mathbb{R}^{+4e} , \mathbf{f} = \mathbf{V} \bm{\beta} \qquad 
\end{equation}
where $\mathbf{V} = \diag{ \{\mathbf{V}_1, \dots, \mathbf{V}_e\} }$, and $\mathbf{f} = [\mathbf{f}_0,...,\mathbf{f}_e]^T$.

Additional constraints on the forces result from Newton-Euler equations:

\begin{align}
\underbrace{\mat{m (\ddot{\mathbf{c}} - \mathbf{g})  \\ m \mathbf{c} \times (\ddot{\mathbf{c}} - \mathbf{g}) + \dot{\mathbf{L}}}}_\mathbf{w}
= 
\underbrace{
\mat{\mathbf{I}_3 & \dots & \mathbf{I}_3 \\
\hat{\mathbf{p}}_1 & \dots & \hat{\mathbf{p}}_e} \mathbf{V}
}_\mathbf{G}
\bm{\beta},
\end{align}
where $\mathbf{w}\in \Rv{6}$ is the so-called gravito-inertial wrench (GIW) \citep{qiu:dhm:2011, Caron2015} and $\hat{\mathbf{p}} \in \R{3}{3}$ is the cross-product matrix associated to $\mathbf{p}$.

Static balance assumes zero acceleration thus we can write:
\begin{align} \label{eq:new_eul}
\mathbf{w}_0 = 
\underbrace{\mat{\mathbf{0}_{3\times 3} \\ m \hat{\mathbf{g}}}}_{\mathbf{D}} \mathbf{c} + 
\underbrace{\mat{-m\mathbf{g} \\ \mathbf{0}}}_{\mathbf{d}}
\end{align}
The wrench $\mathbf{w}_0$ is independent of $c^z$ because of the cross product matrix $\hat{\mathbf{g}}$ inside $\mathbf{D}$, so we can rewrite \eqref{eq:new_eul} as:
\begin{align} \label{eq:new_eul2d}
\mathbf{w}_0 = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d}
\end{align}

If there exists a $\bm{\beta}^*$ satisfying equations \ref{eq:gen} and \ref{eq:new_eul2d}, it means that the configuration is statically balanced.
The problem can then be formulated as an LP:

\begin{equation} \label{eq:lin_prog} \begin{aligned}
\find \quad & \bm{\beta} \in \Rv{+4e} \\
\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
& \bm{\beta} \ge 0 \\
\end{aligned} \end{equation}

\subsubsection{Formulation of a robust LP:}
Let $b_0 \in \mathbb{R}$ a scalar value. We now define the following LP:

\begin{equation} \label{eq:lin_prog_rob} \begin{aligned}
\find \quad & \bm{\beta} \in \Rv{+4e}, b_0 \in \Rv{} \\
\minimize  \quad & -b_0 \\
\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
& \bm{\beta} \ge b_0 \\
\end{aligned} \end{equation}

We observe that if $b_0$ is positive, LP~\ref{eq:lin_prog} admits a solution. The higher $b_0$, the further away from the cone limits admissible forces can be found. If $b_0$ is negative, the configuration is not balanced, and the value $b_0$ indicates "how far" from static balance
the configuration is. $b_0$ can thus be seen as a measure of robustness.

In our implementation, we solve the dual of LP~\ref{eq:lin_prog_rob}.
after a trivial elimination of Lagrange multipliers the dual problem has lower dimensionality:
\begin{equation} \label{eq:dual} \begin{aligned}
\find \quad & \bm{\nu} \in \Rv{6}\\
\minimize  \quad & (\mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d})^T \nu \\
\st \quad &\mathbf{G}^T \bm{\nu} \ge 0 \\
& \mathbf{1}^T \mathbf{G}^T \bm{\nu} = 1 \\
\end{aligned} \end{equation}

Indeed, from Slater's condition, we know that the optimal values of a LP and its dual are equal. Thus the optimal value of LP~\ref{eq:dual} is indeed the optimal $b_0$.

In conclusion, LP~\ref{eq:dual} provides a method to assert rapidly quasi static balance, and returns a robustness value, that can be used as a heuristic for robust static balance.
Another option to enforce robustness is to set a minimum acceptable value for $b_0$; this avoids ponderating balance robustness with other heuristics,
but as a result this conservative approach is not complete.

### 2015-12-04 body/intro.tex
robust equilibrium

The present paper is an extension of a conference paper to appear in the proceedings of the ISRR '15 conference TODO CITE.
The conference paper focuses on the theoritical formulation of the problem, and presents results obtained with virtual avatars.
This extension completes this work with an argumented discussion on the implementation of this theory to real world robots and problems.
In particular, we detail the robust balance criteria designed to ensure the equilibrium of the robot and provide details on the hpp implementation
of our contact generation algorithm. Furthermore, our experiments now concern real robots, namely HRP-2 and Hyq. TODO Appendix describes the generation process
for the robot.

>>>>

The present paper is an extension of a conference paper to appear in the proceedings of the ISRR '15 conference TODO CITE.
The conference paper focuses on the theoritical formulation of the problem, and presents results obtained with virtual avatars.
This extension completes this work with an argumented discussion on the implementation of this theory to real world robots and problems.
In particular, we introduce a robust balance criterion, designed to ensure the equilibrium of the robot. We also provide details on the hpp implementation
of our contact generation algorithm. Furthermore, our experiments are now applied to real robot models, namely HRP-2 and Hyq. TODO Appendix describes the generation process
for the robot.
